\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-13 doc/publications/plos16/plos16.tex
round of changes to plos16.tex

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}.
These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
space where, for example, whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
human genome.

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  
Complex trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
parallelization the split for each node in a tree thereby handling millions of features, such as in the 1000 Genomes
Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
utilizing Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
existing tools even on small datasets where multi-threading generally performs well. Harnessing the virtually unlimited
capability to parallelize tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
variable importance selection~\cite{Szymczak2016}.

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1~and~2), 
\marginpar{something wrong here. We do not report on chromosome 1 and 2.
Aidan} 
the time taken is excessive due to
the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in
\cursedforest\ are an inherent property of random forests. Here we provide demonstrations using synthetic data.  We then
demonstrate the scalability of \cursedforest\ in respect to the dimensionality of the data, building a random forest
model on whole-genome data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) 
and the error.  Finally, given the role different parameter values can play in model construction, we explore the effect tuning
these parameters can have on the prediction accuracy of the model.

@@

Random forests  apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data sets are independently drawn, with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
 $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and  a tree is grown on each of these $B$ samples.
The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}

@@

The benefits of random forest models are:
  \begin{itemize}
  \item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

@@

One of the restrictions preventing Spark ML algorithms from scaling to high-dimensionality data is that the feature-set
for each sample is stored as a vector data type. Spark vectors are stored in resilient distributed dataset (RDDs), and
while RDDs can be split, vectors cannot. So as we add more samples (in our case individuals), Spark ML can simply split
the RDD into more partitions and resource requirements will remain relatively stable. However, as the number of
dimensions (variants) increases, the size of the vectors increases. This results in higher memory-requirements as
increasingly large vectors must be loaded into memory.

@@

The current implementation of \cursedforest\ uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labelled with value $q$ where $q = 1,
  \ldots, Q$ at a node. The entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding \cite{Goldstein.et.al.2011} note that of
the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

\subsubsection{Synthetic data} 
\marginpar{check details of simulation. Piotr}
Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal variables with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
Bernoulli trial with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a very poor fit and still have a useful ranking of variable
  importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
\marginpar{replicated in methods section}
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple additive features.

@@

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/variable_importance.png} 
    \caption{The sum of the variable importance of the 5 significant variables.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/numnber_of_variables_found.png} 
    \caption{The number of the 5 significant variables found in the top 5.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/rbo.png}
    \caption{The  rank-biased-overlap measure.} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
    as a function of \mtry and ntree. {\bf a)} The sum of the variable importances for the 5 significant variables.  {\bf
      b)} OOB (out of bag) classification error. {\bf c)} the number of significant variables found in the top 5. 
     {\bf d)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
  \label{figure:synth}
\end{adjustwidth}
\end{figure}

>>>>

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of life~\cite{Loebbecke2015}.
These data sets are not only growing vertically, by capturing more events, but also horizontally by capturing
more information about these events.  The challenge of big and ``wide'' data is especially pronounced in the health
space where, for example, whole genome sequence (WGS) technology enables researchers to interrogate all 3 billion base pairs of the
human genome.

Identifying the relevant or disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  \marginpar{do  these two sentences say teh same thing?}
Complex trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes
Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
utilizing Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
existing tools even on small datasets where multi-threading generally performs well. Harnessing the virtually unlimited
capability to parallelize tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
variable importance selection~\cite{Szymczak2016}.

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1~and~2), 
\marginpar{something wrong here. We do not report on chromosome 1 and 2.
Aidan} 
the time taken is excessive due to
the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
the model (see Result~\ref{comp}). 
\marginpar{missing ref}
 This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

In the first section we explore some of the properties of random forests as these will also be apparent in the \cursedforest\
implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of
\cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.  Finally, given the role
different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
prediction accuracy of the model.

@@

Random forests  apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data sets are independently drawn, with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
 $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and  a tree is grown on each of these $B$ samples.
The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}

@@

The benefits of random forest models are:
  \begin{itemize}
  \item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees are split on random subsets of variables).
  \end{itemize}

@@

One of the restrictions preventing Spark ML algorithms from scaling to high-dimensionality data is that the feature-set
for each sample is stored as a vector data type. Spark vectors are stored in resilient distributed dataset (RDDs), and
while RDDs can be split, vectors cannot. So as we add more samples (in our case individuals), Spark ML can simply split
the RDD into more partitions and resource requirements will remain relatively stable. However, as the number of
dimensions (variants) increases, the size of the vectors increases. This results in higher memory-requirements as
increasingly large vectors must be loaded into memory.
\marginpar{not clear. Are RDDs the problem or the solutuion}

@@

The current implementation of \cursedforest\ uses an ``information gained'' criteria for splitting. Let $f_q$ be the
fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is $H = -
\sum^{m}_{i=1} f_i \log^{}_2 f_i.$ The expected information gain is the change in entropy $H$ from a prior state $T$ to
a state that takes some information $a$ as given:
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children). 
Variable importance is the sum of IG over all notes and all trees for that variable.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations. 
\marginpar{are they stored a double or integer?}
 For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

\subsubsection{Synthetic data} 
\marginpar{check details of simulation. Piotr}
Each dataset consists of $n$ samples and $p$ variables where $n << p$, and values for each variable are ordinal variables with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
Bernoulli trial with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 1.6, a_{100} = 1.8, a_{1000} = 2.2, a_{10000} = 2.0, a_{10000}=2.3$.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a very poor fit and still have an accurate ranking of variable
  importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.

@@

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/rbo.png}
    \caption{The  rank-biased-overlap measure.} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
    as a function of \mtry and \ntree. {\bf a0} OOB (out of bag) classification error.  {\bf b)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
  \label{figure:synth}
\end{adjustwidth}
\end{figure}
